---
title: Illustration of how Gerlach's method works for data with cluster structure
output:
  word_document: default
  html_document:
    df_data_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning = FALSE, cache = TRUE)
```

These R scripts perfom simulations ("Case with three clusters") in Katahira et al. (Distribution of personality: types or skew?; Commentary: A robust data-driven approach identifies four personality types across four large data sets).

## Preparation

Install the required package (if not yet) by the following commands. 

```{r eval=FALSE}
install.packages("tidyverse")
install.packages("sn")
install.packages("reticulate")
install.packages("mixtools")
install.packages("ks")
install.packages("fields")
install.packages("mvtnorm")
```

Also, please install Python to use scikit-learn library via reticulate library. 
(We recommend Anaconda dibribution  https://www.anaconda.com/ , which contains scikit-learn library.) 

ref: 
https://rstudio.github.io/reticulate/  

## Load library

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
library(reticulate)
library(sn)
library(mixtools) 
library(ks)     # for kernel density estimation
library(fields) # for image.plot
library(mvtnorm)
```

# Case with three clusters

## Synthesize data

Samples are drawn from GMM with three components. 

```{r}
set.seed(1)

n.sample <- 100000  # number of samples 

# true parameters of GMM
true_means <- list(c(0,2), 
                   c(-1.5,-0.5), 
                   c(1.5,-0.5)
                   )
true_sigmas <- list(diag(2)*0.5,
                    diag(2)*0.5,
                    diag(2)*0.5
                    )
true_pi <- c(0.4,0.3,0.3)

dat <- matrix(0,nrow = n.sample, ncol = 2) 
for (idx in 1:n.sample) {
  idxk <- which.max(rmultinom(1,1,true_pi))
  dat[idx,] <- rmvnorm(n = 1,
                      mean = true_means[[idxk]],
                      sigma = true_sigmas[[idxk]]  )
}

df_data <- data.frame(scale(dat))
```

Plot samples and marginals distribution.
```{r}
nplot <- 5000 # number of samples to plot

par(pty = "s")
plot(c(-5,6),c(-5,6),type="n",ann=F)
points(df_data$X1[1:nplot],
       df_data$X2[1:nplot],
       pch=".",cex = 2)

d1 <- density(df_data$X1,from = -4, to = 3)
d2 <- density(df_data$X1,from = -4, to = 3)
lines(d1$x, d1$y*3+4,
      xlim=c(-6,6), ylim=c(0,0.3),lwd = 2)
lines(d2$y*3+4, d2$x,
      xlim=c(-6,6), ylim=c(0,0.3),lwd = 2)
```


Plot shuffuled data.

```{r}
N <- n.sample
xdim <- ncol(df_data)
x.shuffeled <- matrix(0,N,xdim)

for (idx in 1:xdim)
  x.shuffeled[,idx] <- df_data[sample(N),idx]

par(pty = "s")
plot(c(-5,6),c(-5,6),type="n",ann=F) 
points(x.shuffeled[1:nplot,1],
       x.shuffeled[1:nplot,2],
       pch=".",cex = 2)

d1 <- density(x.shuffeled[,1],from = -4, to = 3)
d2 <- density(x.shuffeled[,2],from = -4, to = 3)
lines(d1$x, d1$y*3+4, 
      xlim=c(-6,6), ylim=c(0,0.3),lwd = 2)
lines(d2$y*3+4, d2$x, 
      xlim=c(-6,6), ylim=c(0,0.3),lwd = 2)
```

## Fitting GMMs

Following Gerlach et al., we choose initial parameters of GMMs from the results of K-means. 
If you select "random" initialization, the estimates will be different, suggesting that the GMM estimation for data of this kind is not robust.  

```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
sk <- reticulate::import(module = "sklearn")

klist <- 1:20 # number of components
n.rep <- 10   # number of runs from different intialization

biclog <- numeric(length(klist)) # to store BIC
gmm_list <- list()               # to store GMM

for (idxk in klist){
  K <- klist[idxk]
  ll.min <- -Inf
  
  cat("\nfitting GMM... K =", K)

  for (idxrun in 1:n.rep) {
    cat("*")

    initpar <- "kmeans"
    
    ## uncomment if you want to include random intialization
    # if (idxrun < n.rep*0.5)
    #   initpar <- "kmeans"
    # else 
    #   initpar <- "random"
    
    sk_gmm <- sk$mixture$GaussianMixture
    gmm <- sk_gmm(as.integer(K),
                  n_init = 1L,
                  max_iter = 200L, 
                  init_params = initpar) 
    gmm$fit(df_data)
    
    ll <- gmm$lower_bound_
    
    if (ll > ll.min) {
      ll.min <- ll
      cluster.center <- data.frame(gmm$means_)
      names(cluster.center) <- names(df_data)
      biclog[idxk] <- gmm$bic(df_data)
      gmm_list[[idxk]] <- gmm
    }
  }
}
```

### Plot the locations of Gaussian components

Define a function for plotting cluster centers.
```{r}
plot.cluster.center <- function(cluster.center) {
  
  n.cluster <- nrow(cluster.center)
  for (idxc in 1:n.cluster) {
    points(x = cluster.center[idxc,1],
           y = cluster.center[idxc,2],
           pch = 4, cex = 1.5, col="red", lwd=2)
  }
}
```

Scatter plot of data with ellipses of Gaussian components.

```{r}
par(mfrow=c(3,3), pty = "s",
    mar=c(3.5, 3.5, 2, 1), mgp=c(2.4, 0.8, 0))

for (idxk in 1:8) {
  plot(df_data$X1[1:nplot],
       df_data$X2[1:nplot], "p", 
       pch = ".", 
       xlab="Dimension 1", 
       ylab="Dimension 2", 
       main=paste("K:", idxk))
  
  for (idxc in 1:idxk) {
    mixtools::ellipse(mu=gmm_list[[idxk]]$means_[idxc,], 
            sigma=gmm_list[[idxk]]$covariances_[idxc,,],
            alpha = 1-.6827,  # 1 sigma  
            npoints = 250, col="red") 
  }
  plot.cluster.center(gmm_list[[idxk]]$means_)
}
```


### Plot BIC

```{r warning=FALSE}
par(mgp=c(2, 1, 0), pty = "s")
plot(klist,biclog, type = "o", 
     xlab = "Number of clusters", ylab = "BIC", lwd = 2,
     cex      = 1.5,
     cex.lab  = 1.2,
     cex.axis = 1.2)
```


We select the GMM with three components (K=3). 

```{r}
K.selected <- 3
```

## Evaluate clusters 

### Density estimation
Define a function 'estkd' for density estimation of original data and shuffled data 
```{r}

estkd <- function(x, xmin, xmax, 
                  n.shuffle = 1000, 
                  p.threshold = 0.01, 
                  density.threhold = 0.05,
                  enrichment.threshold = 1.25) {
  
  xdim <- ncol(x) # number of latent factors
  N <- nrow(x)
  
  # kernel density estiamtion of original data
  
  Hpi <- ks::Hpi.diag(x = x, nstage = 2)
  
  k <- ks::kde(x = x,verbose = TRUE, H = Hpi, 
               xmin = xmin, 
               xmax = xmax)
  density.original <- k$estimate
  
  # exceedance of original density
  count_exceedance <- array(0, dim(density.original))
  
  # shuffled data
  cat("Density estimation for shuffled data...")
  cat("\n 1/",n.shuffle, "\n")
  
  for (idxs in 1:n.shuffle) {
    if (idxs %% 100 == 0)
      cat(idxs, " ")
    
    fsc.s.shuffeled <- matrix(0,N,xdim)
    
    for (idx in 1:xdim)
      fsc.s.shuffeled[,idx] <- x[sample(nrow(x)),idx]
    
    k.shuffle <- ks::kde(x = fsc.s.shuffeled, H = Hpi, 
                         xmin = xmin, 
                         xmax = xmax)
    
    count_exceedance <- count_exceedance + 
      as.numeric(density.original < k.shuffle$estimate)
    
    if (idxs == 1)
      density.sum <- k.shuffle$estimate 
    else
      density.sum <- density.sum + k.shuffle$estimate
  }
  
  d.shuffle <- density.sum / n.shuffle
  
  p.value <- count_exceedance / n.shuffle
  sig.region <- (p.value < p.threshold & 
                   density.original > density.threhold & 
                   density.original / d.shuffle > enrichment.threshold)
  
  storage.mode(sig.region) <- "numeric"
  
  list(k = k, # output of kde for original data
       d.original = density.original, 
       count.exceedance = count_exceedance, 
       d.shuffle = d.shuffle,
       p.value = p.value,
       sig.region = sig.region)
}
```

Prepare for plot. 
```{r}
xmin <- c(min(-5,min(df_data)),min(-5,min(df_data)))
xmax <- c(max(5,max(df_data)),max(4,max(df_data))) 

# calculate density
ret <- estkd(df_data, xmin, xmax, 
             n.shuffle = 1000, 
             density.threhold = 0)

zlim_max <- max(ret$d.original)
zlim_min <- 0
cex.main <- 1.8

kd <- ret$k
```

### Scatter plot of original data and density plot
```{r}
par(mfrow=c(1,2), pty = "s")
par(mar=c(3.5, 3.5, 2, 1), mgp=c(2.4, 0.8, 0))

xlim <- c(-4, 4)
ylim <- c(-4, 4)

# Panel 1 (density)
image(kd$estimate,
      x = kd$eval.points[[1]],
      y = kd$eval.points[[2]],
      xlab = "Dimension 1", 
      ylab = "Dimension 2",
      xlim = xlim,
      ylim = ylim, 
      col = viridis::viridis(10),
      cex.axis = 1, 
      cex.lab = 1.2)

# Panel 2 (scatter plot)
plot(c(-4,4),c(-4,4),type="n",
     xlab = "Dimension 1", 
     ylab = "Dimension 2",
     xlim = xlim,
     ylim = ylim,  
     cex.axis = 1, 
     cex.lab = 1.2) 

points(df_data$X1[1:nplot],
       df_data$X2[1:nplot],
       col = rgb(0.2,0.2,0.2,alpha=0.7), 
       pch=".",
       cex = 2)
```

### Plot densities of original data and shuffled data

```{r}
par(mfrow=c(1,2), pty = "s")
par(mar=c(3.5, 3.5, 2, 1), mgp=c(2.4, 0.8, 0))

log.density.original <- log(pmax(kd$estimate,   0.00001))
log.density.shuffle <-  log(pmax(ret$d.shuffle, 0.00001))

# Panel 1 (original data)
image(log.density.original,
      x = kd$eval.points[[1]],
      y = kd$eval.points[[2]],
      xlab = "Dimension 1", ylab = "Dimension 2",
      col = viridis::viridis(20),
      cex.axis = 1.2, cex.lab = 1.5,
      cex.sub = 1.2)

plot.cluster.center(gmm_list[[K.selected]]$means_)

title(main = sprintf("Original data"), 
      cex = 1.2, cex.main = cex.main)

# Panel 2 (shuffled data)
image(log.density.shuffle,
      x = kd$eval.points[[1]],
      y = kd$eval.points[[2]],
      xlab = "Dimension 1", ylab = "Dimension 2",
      col = viridis::viridis(20),
      cex.axis = 1.2, cex.lab = 1.5,
      cex.sub = 1.2)
title(main = sprintf("Null model"),
      cex = 1.5, cex.main = cex.main)
```


### Plot enrichment

```{r}
color.palette = colorRampPalette(
  c("#0080ff","white","#ff8000"))
er <- exp(log.density.original - log.density.shuffle)
par(pty="s")
image.plot(kd$eval.points[[1]],
           kd$eval.points[[2]],
           pmax(pmin(er,1.5),0.5),
           col = color.palette(30),
           zlim = c(0.5,1.5),
           xlab="Factor 1", ylab="Factor 2",
           cex.axis = 1.2, cex.lab = 1.2,cex.sub = 1.2
           )
plot.cluster.center(gmm_list[[K.selected]]$means_)
title(main = sprintf("Enrichment"),
      cex = 1.8, cex.main = cex.main)
```


### p-value and enrichment thresholded
```{r}
par(pty="s")
# p-value (p < .01)
image(kd$eval.points[[1]],
      kd$eval.points[[2]],
      z = (ret$p.value < .01), 
      col = c( rgb(0, 0, 0, alpha=0.0), 
               rgb(0.2,0.2, 0.2, alpha=0.5)),
      xlim = c(-5,5),
      ylim = c(-5,4.5),
      zlim = c(0,1),
      xlab = "Dimension 1", ylab = "Dimension 2",
      cex.axis = 1.5, cex.lab = 1.5
)
# enrichment (> 1.25)
image(kd$eval.points[[1]],
           kd$eval.points[[2]],
           z = exp(log.density.original - log.density.shuffle) > 1.25, 
           col = c( rgb(0, 0, 0, alpha=0.0), 
                    rgb(0, 0, 0.8, alpha=0.5)), 
           zlim = c(0,1),
      add = T
)
plot.cluster.center(gmm_list[[K.selected]]$means_)
```



### Plot 3d density of GMM and and scatter disribution with GMM contours

```{r message=TRUE, warning=FALSE, paged.print=TRUE}
par(mfrow=c(1,2), pty = "s")
par(mar=c(3.5, 3.5, 2, 1), mgp=c(2.4, 0.8, 0))

x <- seq(-4,4, by =0.2)
y <- seq(-4,4, by =0.2)
grid <- expand.grid(x1 = x, x2 = y)

d <- matrix(0, nrow = length(x),
            ncol = length(y))

for (idxc in 1:klist[K.selected]) {
  d <- d + gmm_list[[K.selected]]$weights_[idxc] * 
    matrix(dmvnorm(
      cbind(grid$x1, grid$x2),
      mean = gmm_list[[K.selected]]$means_[idxc,],  
      sigma = gmm_list[[K.selected]]$covariances_[idxc,,]),
      nrow = length(x),
      ncol = length(y))
}
persp(x, y, d, 
      theta=30, 
      phi=20, expand=0.5, 
      ticktype="detailed",
      ltheta = 120, shade = 0.75,
      xlab = "Dimension 1",
      ylab = "Dimension 2",
      xlim = xlim,
      ylim = ylim, 
      zlab = "", 
      cex.axis = 0.8 
      )  

plot(df_data$X1[1:nplot],
     df_data$X2[1:nplot],
     "p",
     col = rgb(0.2,0.2,0.2,alpha=0.7), 
     pch=".",
     xlim = xlim,
     ylim = ylim, 
     xlab="Dimension 1", 
     ylab="Dimension 2",
     cex = 2, 
     cex.axis = 1, 
     cex.lab = 1.2)

for (idxc in 1:klist[K.selected]) {
  mixtools::ellipse(mu=gmm_list[[K.selected]]$means_[idxc,], 
                    sigma=gmm_list[[K.selected]]$covariances_[idxc,,],
                    alpha = 1-.6827, # 1 sigma  
                    npoints = 250, col="red", lwd =2) 
}
plot.cluster.center(gmm_list[[K.selected]]$means_)
```


# Evaluate each cluter center

Define function for evaluate each cluter center
```{r message=TRUE, warning=FALSE, paged.print=TRUE}
# Input：
#    x, # data matrix (factor scores) (respondent x dimension)
#    c.center, cluster center matrix (cluster center x dimension)
#    n.shuffle = 1000  number of shuffles
#
# Output: 
#    $d.original # density of original data 
#    $d.null     # mean density of null model
#    $p.value    # p-value
#    $enrichment # = d.original/d.null
eval_component <- function(x, c.center, n.shuffle = 1000) {
  
  xdim <- ncol(x) 
  N <- nrow(x) 
  n.component <- nrow(c.center)
  
  p.value <- numeric(n.component)
  enrichment <- numeric(n.component)
  
  cat("Bandwidth selection...\n")
  Hpi <- ks::Hpi.diag(x = x[1:min(N,1000),], 
                      nstage = 2)
  cat("kernel dinsity estimation for original data...\n")
  k <- ks::kde(x = x, H = Hpi,
               eval.points = c.center,
               binned = FALSE) 
  density.original <- k$estimate
  
  density.shuffle <- matrix(0, n.shuffle, nrow(c.center))
  
  cat("kernel dinsity estimation for shuffled data...\n")
  for (idxs in 1:n.shuffle) {
    if (idxs %% 10 == 1) 
      cat("=")
    
    sc_shuffeled <- apply(x, MARGIN = 2, sample)
    
    k <- ks::kde(x = sc_shuffeled, H = Hpi, 
                 eval.points = c.center,
                 binned = TRUE) # should be FALSE if the number of dimension >= 4
    
    density.shuffle[idxs,] <- k$estimate
  }
  cat("\n")
  
  density.null <- colMeans(density.shuffle)
  
  for (idx in 1:n.component) {
    
    p.value[idx] <- sum(density.original[idx] < density.shuffle[,idx]) / n.shuffle
    
    enrichment[idx] <- density.original[idx] /
      density.null[idx]
    
    d.null.mean <- mean(density.shuffle[,idx])
  }
  
  list(d.original = density.original,
       d.null = density.null, 
       p.value = p.value,
       enrichment = enrichment )
}
```

Evaluate clusters (Gaussian components). 

```{r message=TRUE, warning=FALSE, paged.print=TRUE}
res_ec <- eval_component(df_data,
               gmm_list[[K.selected]]$means_)

# FALSE: not meaningful, TRUE: meaningful
flg_meaningful <- (res_ec$p.value < 0.01 & 
                     res_ec$enrichment > 1.25) 

# 1: not meaningful,2: meaningful
idx_meaningful <- flg_meaningful + 1 

```


Calculate the fraction of samples classified into each cluster. 

```{r message=TRUE, warning=FALSE, paged.print=TRUE}
cluster_labels <- apply(gmm_list[[K.selected]]$predict_proba(df_data),1,which.max)

(tb <- table(cluster_labels))

cat("Fraction of samples classified into one of the meaningful clusters:",
    sum(tb[flg_meaningful])/sum(tb))
```


